\documentclass[../main.tex]{subfiles}

\begin{document}

\section{Conclusions}\label{sec:conclusions}

I designed and validated a classical attitude control system for an
aerodynamically unstable launch vehicle at the max-$\bar q$ condition, working
entirely in the frequency domain.

\begin{itemize}
\item \textbf{Task 1.} A proportional--derivative attitude law with weak negative
drift feedback stabilises the rigid airframe (open-loop pole at
$+1.84$~rad/s) and meets the design targets $|GM_a|\approx 6$~dB,
$PM\approx 30^\circ$ --- the aerodynamic gain margin and rigid-body phase margin,
read on the full conditionally-stable loop and classified by frequency band.
\item \textbf{Task 2.} Introducing the TVC actuator, the transport delay and the
lightly damped bending mode destabilises the loop through a $+29$~dB resonance.
A four-way filter trade retained a deep gain-stabilising notch at $\omega_{BM}$
(the Eq.-4 lead-lag alone now stabilises the milder peak in $45/75$ cases, but
only marginally, while a notch triplet over-notches and goes unstable at
nominal). Crucially, the actuator, delay and notch lag collapse the rigid phase
margin of the ideal-actuator Task-1 gains from $30^\circ$ to $15^\circ$, so the
PD is \textbf{re-tuned on the full loop} ($K_{D,\theta}: 0.44\to0.69$),
recovering $30^\circ$ and a $165$~ms delay margin with the bending gain-stabilised
at $-18$~dB. The deep notch leans on knowing the bending frequency: it tolerates
$-10\%$ but fails at $+5\%$, whereas a notch$+$lead-lag variant holds the full
$\pm 10\%$ at the price of thinner nominal margins.
\item \textbf{Task 3.} The fixed re-tuned Task~2 controller is robust to
$\pm 30\%$ independent variations of $\mu_\alpha$ and $\mu_c$: all nine cases
(the four box vertices plus the four single-parameter edges) remain stable. The binding limit is the aerodynamic gain
margin (the control-to-aero authority ratio): it erodes as $\mu_\alpha$ grows or
$\mu_c$ shrinks, and the worst vertex V3 (high $\mu_\alpha$ with low $\mu_c$)
compounds both, leaving it at only $0.9$~dB with $18^\circ$ of phase and doubling
the gust excursions --- the quantitative argument for gain scheduling in a flight
design.
\end{itemize}

\paragraph{Limitations and extensions.} Four modelling choices bound the scope
of this design. (i) The pure proportional--derivative law leaves a
steady-state attitude error to a constant gimbal misalignment, which a
proportional--integral--derivative extension would null. (ii) The stiff
second-order TVC model neglects the nozzle back-reaction (the dog-wags-tail
feedback) and the hydraulic load-resonance of the actuator;
retaining the gimbal angle as a state, together with the tail-wags-dog reaction of
Eq.~\eqref{eq:twd}~\cite[Lecture~20]{dclvnotes}, is the natural next fidelity
step. (iii) Sensor placement relative to the bending-mode nodes matters as much
as the filter --- an aft gyro station flips the
sign of the modal slope and can phase-stabilise a mode that would otherwise need
a notch~\cite[Sec.~3.3.2.2]{greensite_cr826} --- whereas here the INS station is
fixed and only the filter is traded. (iv) Propellant sloshing is omitted; it
would add low-frequency, lightly damped modes coupling into the rigid loop and
the drift channel.

The design shows the textbook trade-off of launch-vehicle attitude
control~\cite[Lecture~18]{dclvnotes}: enough gain to stabilise the
unstable airframe, but not so much that it excites the structural modes, and the
Nichols chart lets me see both constraints on one plot. A Simulink
block-diagram mirror of the closed loop reproduces the script results and is
documented in the accompanying \texttt{models/SIMULINK\_GUIDE.md}.

\end{document}
