\documentclass[../main.tex]{subfiles}

\begin{document}

\section{Conclusions}\label{sec:conclusions}

This work transcribed the minimum-fuel powered-descent and pinpoint-landing optimal control problem into a finite-dimensional NLP via trapezoidal direct collocation, and solved it with \texttt{fmincon}/SQP. The main findings are as follows.

\textbf{Direct vs.\ indirect transcription.}\ Compared to the indirect-shooting approach of Homework~1, direct collocation requires no analytical derivation of costate equations or transversality conditions, accepts path constraints (glide-slope, thrust bounds) natively as inequality constraints, and is robust to a coarse initial guess~\cite[Sec.~4.12.7]{betts2010}. The cost is a higher-dimensional NLP ($350$ variables here) and the loss of the structured analytical insight that the bilinear-tangent law provided in the indirect formulation.

\textbf{Max--coast--max thrust structure.}\ The fuel-optimal solution exhibits the classical three-arc bang--off--bang behaviour expected for problems with linear-in-control dynamics and a control-magnitude bound: a saturated braking burn at $T_{\max}$, a midcourse coast arc at $T_{\min} = 0$, and a second saturated burn down to touchdown. This is the direct-transcription analogue of the maximum-principle prediction (Section~\ref{sec:bangbang}): the optimal thrust magnitude never dwells at intermediate values.

\textbf{Sensitivity to flight time.}\ A $\pm 5\%$ sweep around the nominal $t_f = 38$~s revealed monotonic growth of fuel consumption with flight time (a longer horizon stretches the coast arc and increases the cumulative gravity loss), with a $\approx 1.3\%$ propellant variation across the window. The lower bound on $t_f$ is set implicitly by the thrust-magnitude constraint.

\textbf{Active constraints at the optimum.}\ The only path inequality active at the optimum is the upper thrust bound $\|\mathbf{T}\| = T_{\max}$, binding on the two burn arcs with strictly positive KKT multipliers. The glide-slope cone, by contrast, stays narrowly inactive: the trajectory approaches the corridor boundary near the apex of the divert turn with a minimum margin of $1^\circ$--$2^\circ$ across the sensitivity sweep, and the corresponding multipliers are numerically zero --- complementary slackness thus identifies the thrust bound as the single binding constraint shaping the trajectory.

\textbf{ZOH transcription (Task 2).}\ The optional ZOH discretisation of Section~\ref{sec:task2} was implemented in three complementary forms---a nonlinear ZOH with RK4 propagation, the literal Appendix~A construction (LTV-linearised + SCvx) solved by \texttt{fmincon}/SQP, and the same SCvx outer loop with the inner sub-problem cast as a Second-Order Cone Program and solved by \texttt{YALMIP}+\texttt{ECOS}. All three confirm the trapezoidal baseline at the engineering level: final masses agree within a few kilograms ($\lesssim 0.7\%$ of the propellant). The forward-integration test exposes the lower convergence order of the trapezoidal rule, with the ZOH--RK4 variant dropping the per-node fidelity error by five orders of magnitude and sitting close to the \texttt{ode45} tolerance floor. Warm-started from the trapezoidal solution and stabilised by an adaptive trust-region ratio test, both SCvx variants settle on the same nearby local optimum. The conic variant exposes the SOCP structure of the inner step directly and delivers a per-step fidelity that is an order of magnitude tighter than the \texttt{fmincon}/SQP path, demonstrating the convex-sub-problem framework underlying production guidance algorithms (GFOLD, lossless convexification)~\cite{acikmese2007,blackmore2010}.

\textbf{Non-dimensionalisation.}\ Following the same approach used in Homework~1, the OCP is solved in non-dimensional variables (Section~\ref{sec:nondim}), with $L_{ref} = y_0$, $a_{ref} = g$ and the single residual parameter $V_c = V_{ref}/c \approx 0.078$. This shrinks the dynamic range of the decision vector from four orders of magnitude to one and is essential for the LTV linearisation to be well-conditioned: the Jacobians become uniformly $O(1)$ and \texttt{fmincon}/SQP makes appreciable progress per iteration.

\textbf{Limitations and roadmap.}\ The model neglects aerodynamic drag, vehicle attitude dynamics, and any throttling rate limits, and the flight time is fixed. Natural extensions remaining beyond the scope of this report are:
\begin{itemize}
    \item A \textbf{free-time variant} that lifts the fixed-$t_f$ assumption and allows the solver to find the absolute fuel-optimal flight time subject to the active thrust constraint, e.g.\ as in lossless-convexification formulations.
    \item \textbf{Lossless convexification} of a non-zero lower thrust bound $T_{\min} > 0$, which is non-convex in $\mathbf{T}$: with the conic SCvx infrastructure already in place (variant~(c)), the next step is to introduce the slack-variable trick of A\c{c}ikme\c{s}e and Ploen~\cite{acikmese2007} to recover a convex SOCP in the new state, and verify that the original constraint is recovered at the optimum (Section~\ref{sec:lossless}).
    \item The \textbf{virtual-control safeguard} against artificial infeasibility in the SCvx inner problem, which surfaced transiently in Section~\ref{sec:task2}: the penalised buffer $\boldsymbol{\nu}_k$ would make the outer loop robust to cold starts.
    \item Inclusion of \textbf{aerodynamic drag} and \textbf{attitude dynamics}, moving from a 3-DoF point mass to a 6-DoF rigid body with angle-of-attack and angle-of-sideslip dependence. Once the thrust is optimised, the aerodynamic loads also call for a dynamic-pressure constraint $q = \tfrac{1}{2}\rho\,v_{\mathrm{rel}}^2 \leq q_{\max}$ (enforced as a hard path constraint, or checked a posteriori when the thrust profile is prescribed)---the structural max-$q$ load case that is central to Homework~3~\cite[Sec.~4.4.8]{benedikter2022}.
\end{itemize}

\end{document}
